From Freebase to Wikidata: The Great Migration

### 2015-10-17 paper.tex
Addresses Jamie's comments or adds TODOs
Except the last ones about the analysis

Freebase is built on the notions of \emph{objects}, \emph{facts}, \emph{types}, and \emph{properties}.
Each Freebase object has a~stable identifier called a~\emph{``mid''} (for Metaweb ID),
one or more types, and uses properties from these types in order to provide facts.
For example, the Freebase object for Barack Obama has the mid \texttt{/m/02mjmr}
and the type \texttt{/government/us\_president} (among others) that allows the entity to have
a fact with the property \texttt{/government/us\_president/presidency\_number} \linebreak
and the literal integer ``44'' as the value.

@@

The data quality of Freebase was examined thoroughly by the Wikidata community
and anecdotal evidence indeed confirmed that there were certain challenges.
For example, a~small city in France was assigned an ISO country code,%
\footnote{Freebase item page of the city of Saint-Martin: \url{http://www.freebase.com/m/03cc0d0}}
then Boston (the city in Massachusetts) had the type \texttt{/people/person},
and several other occurrences of data quality issues were revealed.
In consequence, a~fully automatic upload of all the content from Freebase into Wikidata
did not seem advisable as the expectations of the Wikidata community regarding the
quality of automatically uploaded data are high (and rightly so).
Despite the fact that an automated upload of all the data
would have led to accelerated growth in the short term,
such a~fast increase of available data would certainly have endangered
the sense of ownership of the Wikidata community for its data.
As an alternative approach, we decided to rely on crowdsourced human curation
and created the \emph{Primary Sources Tool},
a~software solution that displays Freebase statements
for verification to the user that can be added to the currently shown Wikidata item.
The tool will be described in more detail in \autoref{sec:primary-sources-tool}.

@@

Thanks to these efforts, we have already mapped most of the topics that could be mapped automatically,
which is also backed by the comparably small increases gained from the ID reconciliation approach 
and the Knowledge Graph data described in the previous paragraph.
Furthermore, the \emph{mapped} topics have an average number of facts of about~13.9,
whereas the topics that were \emph{not} mapped have an average number of only 5.7~facts.
We have mapped the majority of all topics with more than 47~facts
(191K~non mapped topics \textit{vs.} 195K~mapped topics).
This allows us to conclude that we have mapped the most important items of Freebase.
\autoref{fig:mapped-not-mapped} provides an overview of mapped and non-mapped items
put in relation to the number of facts per item.

@@

For the mapping between Freebase and Wikidata properties, we have chosen to do it by hand
with the help of the Wikidata community.
With that, we have been able to quickly map around 360~properties
(note that Wikidata has a~total of 1,863 properties, and Freebase has around 37,700 properties
\todo{This should probably have a~date.}).
For the mapping of properties where the domain are other topics or literals,
this mapping provides the Wikidata property to use.
We have also implemented a~special case for \texttt{/people/person/parents}
in order to be able to map them to Wikidata properties ``father'' and ``mother''
based on the person gender.
The mapping of CVTs (see \autoref{fig:cvt-obama}) is more complicated,
because it is not possible to have a~\mbox{1-to-1} relationship
between Freebase and Wikidata properties: the CVT is linked to the subject topic by one property
and has properties pointing to its component values, whereas the Wikidata statement
has a~main property value group that is qualified by other such groups.
In order to map a~CVT to a~statement, we have to know which of the CVT's properties
should be used as the main value of the statement, with the others being mapped to qualifiers.
In order to keep a~simple mapping that associates one Wikidata property to one Freebase property,
we map both the property that links the topic to the CVT
and the main property of the CVT to the same Wikidata property, and during the mapping process
pick the latter as the main property of the CVT.
For example, if we map the CVT shown in \autoref{fig:cvt-obama},
we first use the Wikidata property \texttt{position held (P39)} to map
\texttt{/politician/government\_positions\_held} and
\texttt{/government\_position\_held/office\_position\_or\_title}.
Second, for the qualifiers, we use \texttt{start time (P580)}
for \texttt{/government\_position\_held/from} and \texttt{end time (P582)}
for\linebreak \texttt{/government\_position\_held/to}.
With the resulting property mapping, the created Wikidata statement
looks like the one presented in \autoref{fig:statement-obama}.

@@

\subsection{Data Preparation}

In order to prepare the data for integration into the Primary Sources Tool,
we have created a~set of scripts which map the content
of the last Freebase dump to Wikidata statements.%
\footnote{The source code is available under the terms of the Apache~2.0 license:
\url{https://github.com/google/freebase-wikidata-converter}}
The Primary Sources Tool was developed in parallel with the mapping tool.
We first created a~small dataset of around 1M statements
based on few select properties (\emph{e.g.}, \texttt{/people/person/birth\_place})
and deployed both the first dataset and a~basic version of the tool
in order to gather initial user feedback.
In the following \todo{Weeks? Months?}, we progressively added more statements to the tool's database back-end.
Thereby, we were able to slowly upscale the tool's database back-end
without risking back-end performance bottlenecks.
The process of adding data in small batches of 1M~to~2M statements per batch
further allowed us to detect some potential issues in the front-end
and fix them before they became actual problems.
Another positive side-effect of this approach was that it allowed us to gather quick feedback
from the community about the already mapped data and thus to adapt the mapping scripts easily
and correct minor issues with the data early on in the back-end.
Upon approval or rejection of either a~claim or a~reference,
the Wikidata item page in question reloads in order to reflect
the potentially changed state of the migration, as,
in case of incoming claims with references, freshly accepted claims 
can require incoming references to be attached elsewhere in the hierarchy.

@@

Entities and statements are stored in a~relational database (MySQL or SQLite)
and follow the Wikidata data structure with an additional status field
to keep track of statements that either already have been approved or rejected,
or not yet visited.
Statements can be grouped in ``datasets'' and ``uploads'' to
distinguish data from different sources.\footnote{At the time of this writing,
the back-end contained five different datasets.}

@@

\begin{itemize}
  \setlength\itemsep{0em}
  \item 48M~topics
  \item 2,997M triples
  \item 442M facts owned by Google%
    \footnote{That is, the remaining Freebase triples
      after filtering out data about IDs (\texttt{/type/object/id}),
      types (\texttt{/type/object/type}), labels, descriptions (\texttt{/type/object/name},
      \texttt{/common/topic/description}, \emph{etc.}), and triples removed
      for licensing (see \autoref{sec:licensing})}
  \item 68M~labels
\end{itemize}

>>>>

Freebase is built on the notions of \emph{objects}, \emph{facts}, \emph{types}, and \emph{properties}.
Each Freebase object has a~stable identifier called a~\emph{``mid''} (for Machine ID),
one or more types, and uses properties from these types in order to provide facts.
For example, the Freebase object for Barack Obama has the mid \texttt{/m/02mjmr}
and the type \texttt{/government/us\_president} (among others) that allows the entity to have
a fact with the property \texttt{/government/us\_president/presidency\_number} \linebreak
and the literal integer ``44'' as the value.

@@

The data quality of Freebase was examined thoroughly by the Wikidata community
and anecdotal evidence indeed confirmed that there were certain challenges.
For example, a~small city in France was assigned an ISO country code,%
\footnote{Freebase topic page of the city of Saint-Martin: \url{http://www.freebase.com/m/03cc0d0}}
then Boston (the city in Massachusetts) had the type \texttt{/people/person},
and several other occurrences of data quality issues were revealed.
In consequence, a~fully automatic upload of all the content from Freebase into Wikidata
did not seem advisable as the expectations of the Wikidata community regarding the
quality of automatically uploaded data are high (and rightly so).
Despite the fact that an automated upload of all the data
would have led to accelerated growth in the short term,
such a~fast increase of available data would certainly have endangered
the sense of ownership of the Wikidata community for its data.
As an alternative approach, we decided to rely on crowdsourced human curation
and created the \emph{Primary Sources Tool},
a~software solution that displays Freebase statements
for verification to the user that can be added to the currently shown Wikidata item.
The tool will be described in more detail in \autoref{sec:primary-sources-tool}.

@@

Thanks to these efforts, we have already mapped most of the topics that could be mapped automatically,
which is also backed by the comparably small increases gained from the ID reconciliation approach 
and the Knowledge Graph data described in the previous paragraph.
Furthermore, the \emph{mapped} topics have an average number of facts of about~13.9,
whereas the topics that were \emph{not} mapped have an average number of only 5.7~facts.
We have mapped the majority of all topics with more than 47~facts
(191K~non mapped topics \textit{vs.} 195K~mapped topics).
This allows us to conclude that we have mapped the most important items of Freebase.
\autoref{fig:mapped-not-mapped} provides an overview of mapped and non-mapped items
put in relation to the number of facts per item.
\todo{"There is no reference in the text to \autoref{fig:mapped-not-mapped}
(which has a lot of subtle information in it.) It needs some real explanation." -- Jamie}

@@

For the mapping between Freebase and Wikidata properties, we have chosen to do it by hand
with the help of the Wikidata community.
With that, we have been able to quickly map around 360~properties
(note that Wikidata has a~total of 1,863 properties, and Freebase has around 37,700 properties
\todo{This should probably have a~date.}).
For the mapping of properties where the domain are other topics or literals,
this mapping provides the Wikidata property to use.
We have also implemented a~special case for \texttt{/people/person/parents}
in order to be able to map them to Wikidata properties ``father'' and ``mother''
based on the person gender.
The mapping of CVTs (see \autoref{fig:cvt-obama}) is more complicated,
because it is not possible to have a~\mbox{1-to-1} relationship
between Freebase and Wikidata properties: the CVT is linked to the subject topic by one property
and has properties pointing to its component values, whereas the Wikidata statement
has a~main property value group that is qualified by other such groups.
In order to map a~CVT to a~statement, we have to know which of the CVT's properties
should be used as the main value of the statement, with the others being mapped to qualifiers.
In order to keep a~simple mapping that associates one Wikidata property to one Freebase property,
we map both the property that links the topic to the CVT
and the main property of the CVT to the same Wikidata property. Then, during the mapping process
we pick the latter as the main property of the CVT \todo{Rewrite this sentence to be more understandable?}.
For example, if we map the CVT shown in \autoref{fig:cvt-obama},
we first use the Wikidata property \texttt{position held (P39)} to map
\texttt{/politician/government\_positions\_held} and
\texttt{/government\_position\_held/office\_position\_or\_title}.
Second, for the qualifiers, we use \texttt{start time (P580)}
for \texttt{/government\_position\_held/from} and \texttt{end time (P582)}
for\linebreak \texttt{/government\_position\_held/to}.
With the resulting property mapping, the created Wikidata statement
looks like the one presented in \autoref{fig:statement-obama}.

@@

\subsection{Data Preparation}
\todo{"the section seems to describe something more like a phased rollout or release plan than "data preparation"" -- Jamie}

In order to prepare the data for integration into the Primary Sources Tool,
we have created a~set of scripts which map the content
of the last Freebase dump to Wikidata statements.%
\footnote{The source code is available under the terms of the Apache~2.0 license:
\url{https://github.com/google/freebase-wikidata-converter}}
The Primary Sources Tool was developed in parallel with the mapping tool.
We first created a~small dataset of around 1M statements
based on few select properties (\emph{e.g.}, \texttt{/people/person/birth\_place})
and deployed both the first dataset and a~basic version of the tool
in order to gather initial user feedback.
In the following months, we progressively added more statements to the tool's database back-end.
Thereby, we were able to slowly upscale the tool's database back-end
without risking back-end performance bottlenecks.
The process of adding data in small batches of 1M~to~2M statements per batch
further allowed us to detect some potential issues in the front-end
and fix them before they became actual problems.
Another positive side-effect of this approach was that it allowed us to gather quick feedback
from the community about the already mapped data and thus to adapt the mapping scripts easily
and correct minor issues with the data early on in the back-end.
Upon approval or rejection of either a~claim or a~reference,
the Wikidata item page in question reloads in order to reflect
the potentially changed state of the migration, as,
in case of incoming claims with references, freshly accepted claims 
can require incoming references to be attached elsewhere in the hierarchy.

@@

Entities and statements are stored in a~relational database (MySQL or SQLite)
and follow the Wikidata data structure with an additional status field
to keep track of statements that either already have been approved or rejected,
or not yet visited.
Statements can be grouped in ``datasets'' (set of statements from the same origin)
and ``uploads'' (set of statements uploaded at the same time) to
distinguish data from different sources.\footnote{At the time of this writing,
the back-end contained five different datasets.}

@@

\begin{itemize}
  \setlength\itemsep{0em}
  \item 48M~topics
  \item 2,997M triples
  \item 442M facts redistributable by Google\todo{Better formulation?}%
    \footnote{That is, the remaining Freebase triples
      after filtering out data about IDs (\texttt{/type/object/id}),
      types (\texttt{/type/object/type}), labels, descriptions (\texttt{/type/object/name},
      \texttt{/common/topic/description}, \emph{etc.}), and triples removed
      for licensing (see \autoref{sec:licensing})}
  \item 68M~labels
\end{itemize}
